---
title: "Get Data on Languages Spoken in the US"
author: Catherine Brockway, PhD; UN CIFAL Honolulu; catherine.lee.brockway@gmail.com
date: "`r Sys.Date()`"
output: 
  html_document:
    code_download: TRUE
    theme: sandstone
    toc: TRUE
    toc_float: TRUE
    code_folding: hide
---

This project seeks to make data about languages spoken in the US accessible to decision-makers in social sector enterprises. 

```{r env setup, include=FALSE}
library(tidyverse)
library(tidycensus)
library(kableExtra)
library(reactR)
library(huxtable)
#set your census api key using instructions at https://walker-data.com/tidycensus/reference/census_api_key.html
Sys.getenv("CENSUS_API_KEY")
```

Load PUMS Data and Create CSV file
```{r get data, eval=FALSE, include=FALSE}
# lang_pums <- get_pums(
#   variables = c("LANP", "AGEP", "ENG"),
#   state = "all",
#   survey = "acs5",
#   year = 2021,
#   recode = TRUE
# )
# 
# write.csv(lang_pums, "lang_pums.csv")]
# 
# lang_pums <- read_csv("data/lang_pums.csv")
```


Load ACS Data and Create CSV file
```{r get data, eval=FALSE, include=FALSE}
# #view variables in data set
# vars <- load_variables(2021, "acs5", cache = TRUE)
# view(vars)
# 
# #access data and create csv file
# state_pops <- get_acs(geography = "state", 
#                       variables = "B01003_001", #TOTAL POPULATION
#                       year = 2021)
# 
# write.csv(state_pops, "data/state_pops.csv")
```

Read in Data
```{r}
#read in language data
lang_pums <- read_csv("lang_pums.csv")

#read in population data
state_pops <- read_csv("state_pops.csv")
```

Select Columns From State Pops Data Set 
```{r}
state_pops <- state_pops %>%
  rename(state = NAME, pop_estimate = estimate) %>% 
  select(state, pop_estimate)
```


Arrange Language Data
```{r}
totals <- lang_pums %>%
  count(LANP_label, wt = PWGTP)%>%
  arrange(desc(n))
```


```{r reactable US languages}
# reactable(
# totals,
# searchable = TRUE,
# showPageSizeOptions = TRUE,
# striped = TRUE,
# highlight = TRUE,
# columns = list(
# LANP_label = colDef(name = "Language"),
# n = colDef(name = "Number of Speakers")
# )
# )
```

Select Columns
```{r subset data, eval=FALSE, include=FALSE}
lang_nonEng <- lang_pums %>%
  filter(LANP != "000N")%>%
  select(PWGTP, AGEP, ST_label, ENG_label, LANP_label)
```

Create Age Groups
```{r}
#Create breakpoints for the age groups
breakpoints <- c(4, 17, 64, Inf)  # Assuming age groups: 5-17, 18-64, 65 and above

#Create new column with age groups
lang_nonEng$age_group <- cut(lang_nonEng$AGEP, breaks = breakpoints, labels = c("Youth", "Adult", "Senior"))
```

Create New Columns And Rename Columns
```{r}
#create new column for state abbreviations
lang_nonEng <- lang_nonEng %>% separate(col = ST_label, into = c("state", "state_abb"), sep = "/")

#rename columns
lang_nonEng <- lang_nonEng %>%
  rename(pums_person_weights = PWGTP, age = AGEP, eng_level = ENG_label, household_lang = LANP_label)
```

Create Data Frame: Language Totals 
```{r}
#create new table called lang_totals
lang_totals <- lang_nonEng %>%
  group_by(household_lang) %>% 
mutate(total_speakers = sum(pums_person_weights)) %>% #create a column called total_speakers
  select(household_lang, total_speakers) %>% #select only these two columns
  distinct(household_lang, total_speakers)  #remove duplicate rows

#Add commas to the thousandth place
lang_totals$total_speakers <- format(lang_totals$total_speakers, big.mark = ",", scientific = FALSE)


#write.csv(lang_totals, "data/lang_totals2021.csv")
```

Create Data Frame: Languages By State
```{r}
langs_by_state <- lang_nonEng %>% 
  group_by(household_lang, state) %>%
  mutate(speakers_in_state = sum(pums_person_weights)) %>% 
  select(household_lang, state, speakers_in_state) %>% 
  distinct(household_lang, state, speakers_in_state) #remove duplicate rows
```

Merge language and Population Data
```{r}
langs_by_state <- merge(langs_by_state, state_pops, by="state")
```

Find Percentages
```{r}
options(scipen = 999)

langs_by_state <- langs_by_state %>% 
  mutate(percent_state_pop_speaking = speakers_in_state/pop_estimate) %>%   mutate(across(where(is.numeric), ~ round(., digits = 4))) %>%  #Round percentages to two decimal places
select(state, household_lang, speakers_in_state, percent_state_pop_speaking) #select columns

#Add commas to the thousandth place
langs_by_state$speakers_in_state <- format(langs_by_state$speakers_in_state, big.mark = ",", scientific = FALSE)


#write.csv(langs_by_state, "data/langs_by_state.csv")
```


Create Data Frame: Languages by State and Age Group
```{r}
lang_by_state_age <- lang_nonEng %>%
  group_by(age_group, household_lang, state) %>%
  mutate(total_speakers = sum(pums_person_weights)) %>% #create a column called total_speakers
  select(household_lang, age_group, state, total_speakers) %>%
  distinct(household_lang, age_group, state, total_speakers) #remove duplicate rows

#Add commas to the thousandth place
lang_by_state_age$total_speakers <- format(lang_by_state_age$total_speakers, big.mark = ",", scientific = FALSE)

#write.csv(lang_by_state_age, "data/lang_by_state_age2021.csv")
```



Create Data Frame: Languages by State and English Proficiency
```{r}
lang_by_state_english_level <- lang_nonEng %>%
  group_by(eng_level, household_lang, state) %>%
  mutate(total_speakers = sum(pums_person_weights)) %>% #create a column called total_speakers
  select(household_lang, eng_level, state, total_speakers) %>%
  distinct(household_lang, eng_level, state, total_speakers) #remove duplicate rows

#Add commas to the thousandth place
lang_by_state_english_level$total_speakers <- format(lang_by_state_english_level$total_speakers, big.mark = ",", scientific = FALSE)

#write.csv(lang_by_state_english_level, "data/lang_by_state_english_level2021.csv")
```



